Transfer of Learned Opponent Models in Zero Sum Games

### 2020-10-30 Report/_main.Rmd
draft3

@@

>>>>

---
title             : "Learning to learn in a changing environment"
shorttitle        : "Learning to learn"

author: 
  - name          : "Maarten Speekenbrink"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Department of Experimental Psychology, University College London, 26 Bedford Way, London WC1H 0AP, United Kingdom"
    email         : "m.speekenbrink@ucl.ac.uk"
  - name          : "Akhat Rakishev"
    affiliation   : "1"

author_note: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

keywords          : "keywords"
wordcount         : "X"

bibliography      : ["r-references.bib"]

figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes
mask              : no

class             : "man"
output            : papaja::apa6_pdf
---

```{r load-packages, include = FALSE}
library("papaja")
require(knitr)
```

```{r texreport}
texreport <- function(object,effect.name,...) UseMethod("texreport")
texreport.default <- function(object,effect.name,...) return(NULL)

texreport.summary.merMod <- function(object,effect.name,print.estimate=TRUE) {
  if(effect.name %in% rownames(object$coefficients)) {
    tab <- object$coefficients
    estimate <- tab[effect.name,"Estimate"]
    se <- tab[effect.name,"Std. Error"]
    Tval <- tab[effect.name,"t value"]
    df <- tab[effect.name,"df"]
    pval <- tab[effect.name,"Pr(>|t|)"]
    if(pval < .001) ptext <- "$p < .001$" else {
      if(pval < .05) ptext <- paste("$p = .",strsplit(as.character(round(pval,3)),".",fixed=TRUE)[[1]][2],"$",sep="") else {
        ptext <- paste("$p = .",strsplit(as.character(round(pval,2)),".",fixed=TRUE)[[1]][2],"$",sep="")
      }
    }
    txt <- paste("$b = ",round(estimate,2),"$, $SE = ",round(se,2),"$, $t(",round(df,2),") = ",round(Tval,2),"$, ",sep="")
    txt <- paste(txt,ptext,sep="")
  } else {
    txt <- "error"
  }
  txt
}

texreport.summary.Anova.mlm <- function(object,effect.name,effect.size=TRUE,sphericity.correction=TRUE) {
  if(effect.name %in% rownames(object$sphericity.tests) && sphericity.correction && object$sphericity.test[effect.name,"p-value"] < .05) {
    epsilon <- object$pval.adjustments[effect.name,"HF eps"]
    pval <- object$pval.adjustments[effect.name,"Pr(>F[HF])"]
  } else {
    epsilon <- 1
    pval <- object$univariate.tests[effect.name,"Pr(>F)"]
  }
  df1 <- object$univariate.tests[effect.name,"num Df"]*epsilon
  df2 <- object$univariate.tests[effect.name,"den Df"]*epsilon
  Fval <- object$univariate.tests[effect.name,"F"]
  etasq <- object$univariate.tests[effect.name,"SS"]/(object$univariate.tests[effect.name,"SS"] + object$univariate.tests[effect.name,"Error SS"])
  # generate latex output
  if(pval < .001) ptext <- "$p < .001$" else {
    if(pval < .05) ptext <- paste("$p = .",strsplit(as.character(round(pval,3)),".",fixed=TRUE)[[1]][2],"$",sep="") else {
      ptext <- paste("$p = .",strsplit(as.character(round(pval,2)),".",fixed=TRUE)[[1]][2],"$",sep="")
    }
  }
  txt <- paste("$F(",round(df1,2),",",round(df2,2),") = ",round(Fval,2),"$, ",sep="")
  if(effect.size) txt <- paste(txt,"$\\eta^2_\\text{p} = ",round(etasq,2),"$, ",sep="")
  txt <- paste(txt,ptext,sep="")
  txt
}

texreport.linearHypothesis.mlm <- function(object,effect.size=TRUE) {
  if(ncol(object$SSPH) == 1 & nrow(object$SSPH) == 1 & ncol(object$SSPE) == 1 & nrow(object$SSPE) == 1) {
    df1 <- object$df
    df2 <- object$df.residual
    Fval <- (object$SSPH/df1)/(object$SSPE/df2)
    pval <- 1-pf(Fval,object$df,object$df.residual)
    etasq <- object$SSPH/(object$SSPH + object$SSPE)
    # generate latex output
    if(pval < .001) ptext <- "$p < .001$" else {
      if(pval < .05) ptext <- paste("$p = .",strsplit(as.character(round(pval,3)),".",fixed=TRUE)[[1]][2],"$",sep="") else {
        ptext <- paste("$p = .",strsplit(as.character(round(pval,2)),".",fixed=TRUE)[[1]][2],"$",sep="")
      }
    }
    txt <- paste("$F(",round(df1,2),",",round(df2,2),") = ",round(Fval,2),"$, ",sep="")
    if(effect.size) txt <- paste(txt,"$\\eta^2_\\text{p} = ",round(etasq,2),"$, ",sep="")
    txt <- paste(txt,ptext,sep="")
  } else {
    txt <- "error"
  }
  txt
}

texreport.anova.lme <- function(object,effect.name,effect.size=FALSE) {
  if(effect.name %in% rownames(object)) {
    pval <- object[effect.name,"p-value"]
    if(is.null(pval)) pval <- object[effect.name,"Pr(>F)"]
    df1 <- object[effect.name,"numDF"]
    df2 <- object[effect.name,"denDF"]
    Fval <- object[effect.name,"F-value"]
    #etasq <- object$univariate.tests[effect.name,"SS"]/(object$univariate.tests[effect.name,"SS"] + object$univariate.tests[effect.name,"Error SS"])
    # generate latex output
    if(pval < .001) ptext <- "$p < .001$" else {
      if(pval < .05) ptext <- paste("$p = .",strsplit(as.character(round(pval,3)),".",fixed=TRUE)[[1]][2],"$",sep="") else {
        ptext <- paste("$p = .",strsplit(as.character(round(pval,2)),".",fixed=TRUE)[[1]][2],"$",sep="")
      }
    }
    txt <- paste("$F(",round(df1,2),",",round(df2,2),") = ",round(Fval,2),"$, ",sep="")
    #if(effect.size) txt <- paste(txt,"$\\eta^2_\\text{p} = ",round(etasq,2),"$, ",sep="")
    txt <- paste(txt,ptext,sep="")
    txt
  }
}

texreport.anova <- function(object,effect.name,effect.size=FALSE) {
  if(effect.name %in% rownames(object)) {
    pval <- object[effect.name,"Pr(>F)"]
    df1 <- object[effect.name,"NumDF"]
    df2 <- object[effect.name,"DenDF"]
    Fval <- object[effect.name,"F.value"]
    #etasq <- object$univariate.tests[effect.name,"SS"]/(object$univariate.tests[effect.name,"SS"] + object$univariate.tests[effect.name,"Error SS"])
    # generate latex output
    if(pval < .001) ptext <- "$p < .001$" else {
      if(pval < .05) ptext <- paste("$p = .",strsplit(as.character(round(pval,3)),".",fixed=TRUE)[[1]][2],"$",sep="") else {
        ptext <- paste("$p = .",strsplit(as.character(round(pval,2)),".",fixed=TRUE)[[1]][2],"$",sep="")
      }
    }
    txt <- paste("$F(",round(df1,2),",",round(df2,2),") = ",round(Fval,2),"$, ",sep="")
    #if(effect.size) txt <- paste(txt,"$\\eta^2_\\text{p} = ",round(etasq,2),"$, ",sep="")
    txt <- paste(txt,ptext,sep="")
    txt
  }
}

```{r read-data,warning=FALSE}
source("R/readData.R")
demo_exit <- read.csv("data/demographics_and_exit_clean.csv")
trial_data <- read.csv("data/trial_data_clean.csv")
trial_data$trial <- 1:100
```

# Introduction

## Optimal learning

If the volatility $\sigma_{\xi}(t)$ and noise $\sigma_\epsilon$ are know, optimal Bayesian learning is implemented by the Kalman filter
$$\hat{\mu}_{t+1} = \hat{\mu}_t + k_t (y_t - \hat{\mu}_{t})$$
where the "Kalman gain" is $$k_t = \frac{s_t + \sigma^2_{\xi}(t)}{s_t + \sigma^2_\xi(t) + \sigma^2_\epsilon}$$ and $s_t$ is the prior uncertainty about the mean $\mu_t$, computed recursively as $$s_t = (1-k_t)(s_{t-1} + \sigma^2_\xi(t))$$. Note that the updating of estimates $\hat{\mu}_t$ takes the same form as the well-known delta-rule, where each update is proportional to the prediction error ($y_t - \hat{\mu}_t$), with the proportionality constant $k_t$ is the learning rate, which is usually constant in the delta-rule, but time-varying in the Kalman filter and dependent on the prior uncertainty ($s_t$), and the level of volatility ($\sigma_\xi$) and environmental noise ($\sigma_\epsilon$). If both latter parameters are constant, then as $t \rightarrow \infty$, the Kalman gain converges to a constant and thus becomes equal to the constant-learning rate delta-rule. This is the steady-state Kalman filter, and provides a means to determine the optimal (constant) learning rate. The steady state is 
\begin{equation}k(t)^* = \frac{1}{2} \left(\sqrt{\frac{\sigma^4_\xi}{\sigma^4_\epsilon} + 4 \frac{\sigma^2_\xi}{\sigma^2_\epsilon}} - \frac{\sigma^2_\xi}{\sigma^2_\epsilon} \right)\end{equation}

## A particle filter

When the parameters are unknown, the optimal learning rule becomes intractible. Bayesian scheme's for estimating the parameters on-line need to rely on approximations, such as Monte Carlo Markov Chain estimation, which are impractible for on-line estimation, as for each time point, estimation would have to be started fresh. When the parameters are time-varying,

# Methods
We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study. <!-- 21-word solution (Simmons, Nelson & Simonsohn, 2012; retrieved from http://ssrn.com/abstract=2160588) -->

## Participants and design

One hundred participants (`r  sum(dat$gender[dat$trial == 1] == "Female",na.rm=TRUE)` females, `r  sum(dat$gender[dat$trial == 1] == "Male",na.rm=TRUE)` males, `r sum(is.na(dat$gender[dat$trial == 1]))` other or preferred not to say), aged between `r min(dat$age)` and `r max(dat$age)` ($M = `r round(mean(dat$age[dat$trial==1]),2)`$, $SD = `r round(sd(dat$age[dat$trial==1]),2)`$), voluntarily participated in the study. The participants were recruited from the UCL Psychology and Language Sciences subject pool or by personal invitation from the researchers. Eligible university students ($n=`r sum(dat$version[dat$trial==1] == "sona")`$) received course credit for their participation. In addition, all participants were invited to enter a draw to win one of two gift vouchers (worth £50 and £20). Participants were informed that the chance of winning the prizes was dependent on their performance during the experiment. The study was approved by the UCL ethics committe and all participants provided informed consent before participating in the experiment.

Participants were randomly allocated to either the Gradual change and Abrupt change condition. In the gradual condition, the level of volatility changed according to a sinusoid function, smoothly increasing and decreasing over time. In the abrupt condition, the level of volatility changed in a stepwise manner, being low in some blocks of trials and abruptly increasing in others.

## The Fisher Game

All participants played "The Fisher Game", an online browser-based game programmed in HTML and javascript. In the game, particpants controlled the position of a fisherman's boat in order to catch as many fish as possible. They could maximise the number of fish caught by placing the horizontal position of the boat such that the fishing rod was located at the centre of a school of fish. The further away the rod was from the centre of the school of fish, the less fish they could catch. The horizontal position of the school of fish moved from trial to trial, with the level of movement dependent on the level of volatility. On each of 100 trials participants were asked to choose a position for their boat, effectively predicting were the school of fish would be on that trial. After making their prediction, participants received information about the actual location of the school of fish and how many fish they caught. The game then proceeded to the next trial.

More formally, on each trial $t$ the location of fish $y_t$ was equal to mean fish location $\mu_t$ plus random noise $\epsilon_t$. The mean fish location $\mu_t$ was in turn equal to of the mean fish location on a previous trial, $\mu_{t-1}$,  plus random change $\zeta_t$ representing the effect of volatility:

$$\begin{aligned}
y_t &= \mu_t + \epsilon_t && & \epsilon_t &\sim \mathcal{N}(0,\sigma_\epsilon) \\
\mu_t &= \mu_{t-1} + \xi_t && & \zeta_t &\sim \mathcal{N}(0,\sigma_\xi(t))
\end{aligned}$$

The standard deviation of the noise was set to $\sigma_\epsilon = 4$ in all conditions and did not vary over time. The level of volatility, $\sigma_\xi(t)$ (the standard deviation of the changes in mean fish location) varied over time in a way that was specific to each condition. In the Gradual change condition, the volatility varied according to a sinus function:

$$ \sigma_\xi(t) = 16.5 + 13.5 \times \sin \left( -2.5 + \frac{5 \times (t - 1)}{99} \right)$$

In the Abrupt change condition, volatility varied in stepwise way:

$$\sigma_\xi(t) = \begin{cases} 3 & \text{if } t \in [1;25] \text{ or } [51;75] \\ 30 & \text{if } t \in [26;50] \text{ or } [76,;100] \end{cases} $$
In both conditions lowest volatility was equal to $\sigma_\xi(t) = 3$ and the highest volatility was equal to $\sigma_\xi(t) = 30$. Overall, in both conditions there were two low and two high volatility periods.

```{r task-structure, out.width = "85%", fig.cap = "Data generating process used in the experiment. a) schematic representation of data generation process; b) example of generated data: graphs in the top row show variation of volatility by trial, graphs in the middle row show mean fish location, graphs in the bottom row show observed fish locations."}
include_graphics("img/task_structure.eps")
```

We generated ten different random sequences of fish location per condition in advance according to the data generating process described above. This allowed to let each player observe a randomly generated set of locations, making it unlikely that the results would be dependent on the particulars of a single random sequence or that players would be able to help each other by discussing known fish locations.

The number of fish caught each trial  was then determined by dividing 0.1 by absolute prediction error (difference between response  and actual location , see the formula below). As generated data and participants’ responses were transformed to fit a 0-1 scale, the reward formula allowed players to catch very high amounts of fish in case of very accurate predictions (capped at 300). At the same time, they were still able to catch few fish even in case of very inaccurate predictions. We used this scoring method to keep players motivated, as due to the randomness in the data generating process it was impossible to exactly predict the next fish location. The scores were not used in actual analysis.

Participants completed the game at their own pace in a single session. After providing informed consent, participants provided demographic information (age, gender, and nationality). They then read instructions, in which they were informed about the game interface, the goal of the game, and the conditions of the prize draw.

Following the instructions the game started. On each trial, participants could see the location of the school of fish on the previous trial on the game viewport. They were also able to see all past locations of the school of fish in the "location graph" on the right of the game viewport. They then chose the horizontal position of the fishing boat through a mouseclick at any position on the game viewport. After making their choice, the boat smoothly moved to the chosen location, and the school of fish moved to the new location. Participants were then informed about the number of fish they caught on that trial.

```{r game-screen, out.width = "85%", fig.cap = "Illustration of online game interface. Yellow and blue text labels point on to respective interface elements."}
include_graphics("img/game_screen.eps")
```

Upon completion of the game participants were redirected to a post-game questionnaire were they were asked to answer the following three questions about the study: "How do you think fish locations were determined from day to day?", "What was your strategy to determine where to fish next?", and "What do you think was the aim of this study?". The answers were given in a free format and were not mandatory.

# Results

## Exclusions

Four of the 100 participants were excluded from further analysis because both the mean squared error of their choices and the mean reaction time further than two standard deviations removed from the sample mean. Of the remaining `r with(subset(trial_data,trial == 1),length(condition))` participants `r with(subset(trial_data,trial == 1),{tab <- table(condition); tab[names(tab) == "abrupt"]})` were in the abrupt and `r with(subset(trial_data,trial == 1),{tab <- table(condition); tab[names(tab) == "gradual"]})` in the gradual condition.

## Performance

Participants' average mean squared error was equal to 0.012 (SD = 0.006) and average mean reaction time to 2207.43 (SD = 949.815) milliseconds. Mean reaction times did not differ between conditions (t(72.37) = 1.17, p = .248). Mean squared error was significantly higher (t(78.09) = -4.23, p < .001) in the gradual condition (M = .015) than in the abrupt condition (M = .010).

## Empirical learning rates

According to the "delta-rule", fish locations are learned as

$$\hat{y}_{t+1} = \hat{y}_{t} + \eta(t) \left( y_{t} - \hat{y}_{t} \right)$$
where we allow the learning-rate $\eta(t) > 0$ to vary from trial to trial. Assuming that participants location choices $r_t$ are a noisy reflection of their location predictions:
$$\begin{aligned} r_t = \hat{y}_t + e_t && e_t \sim \mathcal{N}(0,\sigma_e)\end{aligned}$$
we substitute $r_t = \hat{y}_t$ and re-arrange this equation to

$$\begin{aligned} r_{t+1} - r_{t} &= \eta(t) \left( y_{t} - r_{t} \right) \\ u_t &= \eta(t) d_t \end{aligned}$$
with $u_{i,t} = (r_{i,t+1} - r_{i,t})$ representing the update in location choice from trial $t$ to trial $t+1$ and $d_{i,t}= (y_{i,t} - r_{i,t})$ the deviation between location choice and true location on trial $t$. To assess the relation between $\eta(t)$ and the optimal learning rate $\omega_t = \frac{\sigma_\xi^2(t)}{\sigma_\xi^2(t) + \sigma^2_\epsilon}$, we estimated a linear mixed-effects model

<!-- with $u_{i,t} = (r_{i,t+1} - r_{i,t})$ (the update in location choice from trial $t$ to trial $t+1$ for participant $i$) as dependent variable and $d_{i,t}= (y_{i,t} - r_{i,t})$ (the deviation between location choice and true location on trial $t$ for participant $i$) as predictor: -->

$$u_{i,t} = \left(\beta_d + b_{di} + \beta_{dC} C_i + \beta_{d\omega} \omega_{it} + b_{d\omega i} \omega_{it} \right) d_{it} + e_{t,i}$$
where $C$ is a contrast-coded predictor for condition ($C = 1$ for the abrupt, and $C = -1$ for the gradual condition), each $\beta$ term is a fixed effect, and each $b$ term a participant-wise random effect. Random intercepts and slopes were allowed to correlate. The model was estimated with the lme4 package for R [@R-lme4], and $p$-values were computed with the lmerTest package [@R-lmerTest]. 
```{r mixed-effects,cache=TRUE,echo=FALSE}
require(lmerTest)
dat <- subset(trial_data,trial < 100)
dat$omega <- .5*(sqrt(dat$volatility^4/4^4 + 4*dat$volatility^2/4^2) - dat$volatility^2/4^2) # (2*dat$volatility^2)/(2*dat$volatility^2 + 4^2)

mod0 <- lmer(update ~ condition_as_number*omega*error + (scale(omega)*error|id),data=dat)
mod1 <- lmer(update ~ error + condition_as_number:error + omega:error + condition_as_number:omega:error - 1 + (0 + error + scale(omega):error|id),data=dat)
mod1_summ <- summary(mod1)
mod0_summ <- summary(mod0)
aov_full_reduced <- anova(mod0,mod1)
```
There was a significant main effect of deviation, `r texreport(mod1_summ,"error")`. This effect of error was moderated by the steady-state learning rate, `r texreport(mod1_summ,"error:omega")` and condition, `r texreport(mod1_summ,"error:condition_as_number")`. There was also a three-way interaction between deviation, optimal learning rate, and condition, `r texreport(mod1_summ,"error:condition_as_number:omega")`.

We also estimated a linear-mixed effects model which, in addition to these effects, included fixed effects of condition ($C$), optimal learning rate ($\omega$), as well as all two-way and three-way interactions between these variables and deviation ($d$). This model also included an additional random effect for optimal learning rate ($\omega$). Comparing these two models indicated that the latter, more complex model fitted the data significantly better than the first, simpler model, $\chi^2(`r aov_full_reduced$'Chi Df'[2]`) = `r printnum(aov_full_reduced$'Chisq'[2])`$, $p = `r round(aov_full_reduced$'Pr(>Chisq)'[2],3)`$. Hence, we report the results from the full model. There was a significant main effect of deviation, `r texreport(mod0_summ,"error")`. This effect of error was moderated by the optimal learning rate, `r texreport(mod0_summ,"omega:error")` and condition, `r texreport(mod0_summ,"condition_as_number:error")`. There was also a three-way interaction between deviation, optimal learning rate, and condition, `r texreport(mod0_summ,"condition_as_number:omega:error")`. Finally, there was a significant main effect of condition, `r texreport(mod0_summ,"condition_as_number")`, which is difficult to interpret, but indicates a small bias where ...

These effects are more easily interpreted when inspecting the model-predicted learning rate (Figure \@ref(fig:learning-rate-plot)). As can be seen there, while on average, people adapt their learning rate to the changes in volatility, the adaptation is more marked in the gradual compared to the abrupt change condition. Also, the predicted learning rates are mostly higher than the optimal steady-state learning rates.  
```{r learning-rate-plot,echo=FALSE,out.width = "85%", fig.width=6, fig.height=3, fig.cap="Predicted learning rates from the linear mixed-effects model as a function of the optimal learning rate."}
require(ggplot2)
require(lmerTest)
coef <- fixef(mod0)

# newdata <- expand.grid(condition=c("abrupt","gradual"),omega = seq(min(dat$omega),max(dat$omega),length=100))
# newdata$condition_as_number <- 1
# newdata$condition_as_number[newdata$condition == "abrupt"] <- -1
# coeff <-  fixef(mod0)
# 
# newdata$pred_eta <- coeff["error"] + coeff["condition_as_number:error"]*newdata$condition_as_number + coeff["omega:error"]*newdata$omega + coeff["condition_as_number:omega:error"]*newdata$condition_as_number*newdata$omega
# ggplot(newdata,aes(x=omega,y=pred_eta,colour=condition)) + geom_line() + ylab("predicted learning rate") + xlab("optimal learning rate") + ylim(c(min(newdata$omega),max(newdata$pred_eta))) + xlim(c(min(newdata$omega),max(newdata$pred_eta))) + geom_abline(slope=1,intercept=0,lty=2,colour=grey(.5))

newdata <- data.frame(condition=rep(c("abrupt","gradual"),each=100),volatility = c(subset(trial_data,condition=="abrupt")$volatility[1:100],subset(trial_data,condition=="gradual")$volatility[1:100]),trial=1:100)
newdata$omega <- .5*(sqrt(newdata$volatility^4/4^4 + 4*newdata$volatility^2/4^2) - newdata$volatility^2/4^2) # (2*newdata$volatility^2)/(2*newdata$volatility^2 + 4^2)
newdata$condition_as_number <- 1
newdata$condition_as_number[newdata$condition == "abrupt"] <- -1
newdata$type = "optimal"

newdata <- rbind(newdata,cbind(newdata[,1:5],type="estimated"))

coeff <-  fixef(mod1)
newdata$omega[newdata$type=="estimated"] <- coeff["error"] + coeff["error:condition_as_number"]*newdata$condition_as_number[newdata$type=="estimated"] + coeff["error:omega"]*newdata$omega[newdata$type=="estimated"] + coeff["error:condition_as_number:omega"]*newdata$condition_as_number[newdata$type=="estimated"]*newdata$omega[newdata$type=="estimated"]

# coeff <-  fixef(mod0)
# newdata$omega[newdata$type=="estimated"] <- coeff["error"] + coeff["condition_as_number:error"]*newdata$condition_as_number[newdata$type=="estimated"] + coeff["omega:error"]*newdata$omega[newdata$type=="estimated"] + coeff["condition_as_number:omega:error"]*newdata$condition_as_number[newdata$type=="estimated"]*newdata$omega[newdata$type=="estimated"]

ggplot(newdata,aes(x=trial,y=omega,colour=type)) + geom_line() + ylab("(predicted) learning rate") + xlab("trial") + facet_wrap(~condition) + theme(legend.position="bottom") #+ ylim(c(min(newdata$omega),max(newdata$pred_eta))) + xlim(c(min(newdata$omega),max(newdata$pred_eta))) + geom_abline(slope=1,intercept=0,lty=2,colour=grey(.5))

<!--
. However, as the comparison just fails significance, we also report the results of the more complex model in the appendix.

provides us with a trial-by-trial estimate of the learning rate
$$\hat{\eta}(t) = \frac{r_{t + 1} - r_{t}}{y_t - r_t}$$
Under the assumption that $r_t$ is Normal distributed, this empricial learning rate follows a general Cauchy distribution.
-->

# References
```{r create_r-references}
r_refs(file = "r-references.bib")
```

<!--chapter:end:article.Rmd-->

---
title             : "brouillon"
shorttitle        : "Draft-3"

bibliography      : ["Mendeley2.bib"]
csl               : "apa-6th-edition.csl"

documentclass     : "apa6"
# classoption       : a4paper
classoption       : man
output            : papaja::apa6_pdf
citation_package  : natbib
---

```{r load-packages, include = FALSE}
library(papaja)
require(knitr)
require(citr)
require(bookdown)

# using some functions dplyr, ggpubr, PairedData and sjPlot. Need to be loaded. 
library(plyr)
library(tidyr)
library(dplyr)
library(MASS)
library(ggpubr)
library(afex)
library(PairedData)
library(multcompView)
library(lsmeans)
library(magick)
library(depmixS4)
```

# Introduction
Being able to transfer previously acquired knowledge to a new domain is one of the hallmarks of human intelligence. Humans are naturally endowed with the ability to extract relevant features from a situation, identify the presence of these features in a novel setting and use previously acquired knowledge to adapt to previously unseen challenges using acquired knowledge. More formally, @perkins1992transfer defines transfer of learning as the application of skills, knowledge, and/or attitudes that were learned in one situation to another learning situation. This typically human skill has so far eluded modern AI agents. Deep neural networks for instance can do very well on image recognition tasks and can even reach super-human performance levels on video and strategic board games. Yet they struggle to learn as fast or as efficiently as humans do, and more importantly they have a very limited ability to generalize and transfer knowledge to new domains. [@Lake2017] argue that human learning transfer abilities take advantage of important cognitive building blocks such as an abstract representation of concepts underlying tasks and compositionally structured causal models of the environment.

<!--chapter:end:brouillon.Rmd-->

---
title             : "Transfer of Learned Opponent Models in Zero Sum Games"
shorttitle        : "Draft-1"

bibliography      : ["Mendeley.bib"]
# csl               : "apa-6th-edition.csl"

```{r load-packages, include = FALSE}
library(papaja)
require(knitr)
require(citr)
require(bookdown)
# using some functions dplyr, ggpubr, PairedData and sjPlot. Need to be loaded. 
library(tidyr)
library(dplyr)
library(MASS)
library(ggpubr)
library(afex)
library(PairedData)
library(multcompView)
library(lsmeans)
library(magick)
```

Finally, the analysis showed a significant interaction effects of block by game ( F(2,100) = 6.92 , p = .002, $\eta^{2}$ = 0.02), indicating that within-game learning differed between the games. Indeed, second half scores in RPS are significantly higher than first-half scores (t(150) = 5.59, p <.0001), while there was no significant difference between block scores for the other two games. This is indicative of the significant within game transfer within the first game when players have no epxerience against the opponent, as opposed to much lower within game performance improvement in the latter games when participants have had some exprience playing against the opponenet and start with higher scores indicative of transfer.  There was also a three-way interaction between condition, game, and block ( F(2,100) = 3.88 , p = .023, $\eta^{2}$ = 0.01), which indicates, as seen in Figure\ \@ref(fig:exp1-3factor-plot) that within-game learning changes across games also depend on the sophistication of the opponent. For instance, there is more within game learning in the third game against level-2 opponents, since the initial scores are lower than agaisnt level-1 opponent. The explanation for this will become clearer when we discuss the factors moderating learning transfer in the next section.

As a measure for learning transfer, we focus on participants’ scores in rounds 2-6, excluding the very first round for which the agent is programmed to play randomly as it has no data on prior rounds on which to build its response. A group of players with no experience of the game are expected to have scores not significantly different from 0. Any significantly positive group average scores would therefore reflect prior learning from past experiences.

apa_lm1 <- apa_print(exp1_early_anova)

The participants each played 3 games sequentially against both level-1 and level-2 computer opponents, rather than just one like in the first experiment. The three games were Rock-Paper-Scissors, Fire-Water-Grass, and the penalty shootout game. The first two games were identical to the ones used in the first experiment. In the final game (shootout) the participants played the role of the player shooting a football (soccer) penalty, and the AI opponent was the goalkeeper. Players had the choice between three actions, like in the first two games: Shooting the football to the left, right or centre of the goal. If the player shoots in a direction different from that of where the goalkeeper dives, they win the round and the goalkeeper loses. Otherwise, the goalkeeper catches the ball and the player loses the round. There is no possibility of ties in this game. Figure XXXXXXXXX shows a snapshot of play in the shootout game. What makes this game different however is that there are two ways to beat the opponent in each round: if we think the opponent is going to choose ''right'' in the next round, we can win by both choosing ''left'' and ''center''. A level-1 player (thinks that his opponent will repeat his last action) has two ways to win the next round. As such, we have programmed the level-2 computer program to choose randomly between the two possibilities that a level-1 player may choose.

All_comp_results <- read.csv(file="../All_results.csv")
exp1_comp_table <- table(All_comp_results[, "condition"],c("Random","ToM_BT","ToM_NBT","QL", "QL_states","EWA","S_EWA","EWA_states", "MBM")[apply(All_comp_results[,c("Random_BIC","Bayes_Tr_BIC","Bayes_No_Tr_BIC","QL_BIC","QL_states_BIC","EWA_BIC","EWA_self_BIC","EWA_states_BIC")],1,which.min)])

exp1_model_comp <- data.frame()
for(id in unique(exp1_dat$human_id)) {
  tdat <- subset(exp1_dat,human_id == id)
  tot_score <- sum(tdat$score)
  tot_time <- sum(tdat$human_rt)
  early_dat <- subset(tdat,between(tdat$round,2,6) & (game =="fwg"))
  #early_dat <- subset(tdat,between(tdat$round,2,6) & (game =="fwg" | game =="numbers") )
  tr_score <- sum(early_dat$score)
  id_results <- subset(All_comp_results, ID == id)
  min_BIC <- apply(id_results[,c("Random_BIC","Bayes_Tr_BIC","Bayes_No_Tr_BIC","QL_BIC","QL_states_BIC","EWA_BIC","EWA_self_BIC")],1,min)

```{r exp1-comp-models, fig.cap = "Histogram of best fitting computational models by condition for experiment 1", fig.align='center'}

```{r exp1-cumScores, fig.cap = "Average cumulative scores of participants in experiment 1 by best fitting model", fig.align='center'}
p5 <- ggplot(data = dat_by_model, aes(x = t, y=model_acc_sum, group = Best_model)) + 
   geom_line(aes(color= Best_model))

p5+ labs(color = "Best fitting model", x = "Round Number", y="Accumulated score")
```

```{r exp2-comp-models, fig.cap = "Histogram of best fitting computational models by condition for experiment 2", fig.align='center'}

#knitr::include_graphics("../Report/images/exp1_comp_models.png", dpi = 108)
par(las=2) # make label text perpendicular to axis
par(mar=c(5,8,4,2)) # increase y-axis margin.
barplot(exp2_comp_table,
        horiz =TRUE, # rotate barplot for better visibility 
        las=1, # change orientation x axis labels 
        cex.names=0.8, # text label size
        beside =TRUE,
        xlab="Number of participants",
        # legend = c("level-1 opp first","level-2 opp first"),
        legend = paste0(rownames(exp2_comp_table) , " faced first"),
        args.legend=list( # positioning of legend box 
        x=ncol(exp2_comp_table) + 6,
        y=max(colSums(exp2_comp_table)) +1,
        bty = "n")
 )

```{r exp2-cumScores, fig.cap = "Average cumulative scores of participants in experiment 2 by best fitting model", fig.align='center'}

Plotting cumulative scores by best model for experiment 2, we see very similar results looking at Figure\ \@ref(fig:exp2-cumScores), in that participants whose behavior was best fit by a ToM model of learning the oppponent strategy had the highest cumulative performance. Out of these ToM models, the one in which there is within-game but no between-game transfer (NBT) had the best cumulative performance, folowed by a model in which both within and between trasnfer of opponent models is allowed (BT). The next best model from a performance perspective was a Q-learning model with states and within game transfer, followed by ToM models where either players started learning from scratch at each stage of each game (NT) and the model in which they didn't distinguish between opponenets (Naive). As expected, random play as well as self-tuning EWA were at the bottom of cumulative performance.

Our online experiments confirm behavioural game theory results, stating that human players can deviate from Nash equilibrium play and learn to adapt to the opponent strategy and exploit it when the opponent itself is deviating from Nash equilibrium. Moreover, we showed that participants do transfer their learning to new games with varying degrees of similarity. The most likely explanation as we saw is that players build a mental representation of their opponent strategy, rather than rely on associative learning or action-based heuristics. The transfer is also moderated by the level of sophistication of the opponent, with participants showing more success in learning and transferring against opponents adopting a less sophisticated strategy.

Having said that, there remains a high degree of heterogeneity between players. There is a high positive association between players who learn to beat the sophisticated and less sophisticated opponents, indicating that some players are more able to detect the patterns in opponent play and learn how to exploit them. Moreover, the computational modelling shows that a significant number of players are still using random strategies in the games. Whether this ability depends on engagement with the experiment or on other characteristics of the learner is an open question that more research in the area would hope to answer.

```{r create_r-references}
r_refs(file = "Mendeley.bib")

<!--chapter:end:draft_report_v1.Rmd-->

@@

<!--chapter:end:draft_report_v2.Rmd-->

---
title             : "Transfer of Learned Opponent Models in Zero Sum Games"
shorttitle        : "Draft-3"

# author_note: |
  #<!-- Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line. -->

  #Enter author note here.

bibliography      : ["Mendeley2.bib"]
csl               : "apa-6th-edition.csl"

documentclass     : "apa6"
# classoption       : a4paper
classoption       : man
output            : papaja::apa6_pdf
      pandoc_args : !expr rmdfiltr::add_wordcount_filter()
#citation_package  : natbib
---

```{r load-packages, include = FALSE}
library(papaja)
library(kableExtra)
require(knitr)
require(citr)
require(bookdown)

# using some functions dplyr, ggpubr, PairedData and sjPlot. Need to be loaded. 
library(plyr)
library(tidyr)
library(dplyr)
library(MASS)
library(ggpubr)
library(afex)
library(PairedData)
library(multcompView)
library(lsmeans)
library(magick)
library(depmixS4)
library(rmdfiltr)
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)

```{r, include=FALSE} 
options(tinytex.verbose = TRUE)
add_wordcount_filter(NULL, error = FALSE)
```

# Introduction
Being able to transfer previously acquired knowledge to a new domain is one of the hallmarks of human intelligence. Humans are naturally endowed with the ability to extract relevant features from a situation, identify the presence of these features in a novel setting and use previously acquired knowledge to adapt to previously unseen challenges using acquired knowledge. More formally, @perkins1992transfer defines transfer of learning as the application of skills, knowledge, and/or attitudes that were learned in one situation to another learning situation. This typically human skill has so far eluded modern AI agents. Deep neural networks for instance can do very well on image recognition tasks and can even reach super-human performance levels on video and strategic board games. Yet they struggle to learn as fast or as efficiently as humans do, and more importantly they have a very limited ability to generalize and transfer knowledge to new domains. @Lake2017 argue that human learning transfer abilities take advantage of important cognitive building blocks such as an abstract representation of concepts underlying tasks and compositionally structured causal models of the environment.
